---
title: "COVID19 Report"
author: "Konica Patait"
date: "2023-01-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(dplyr)
library(lubridate)
```

# COVID19 Analysis
This report is based on COVID-19 data that is publicly available on [Johns Hopkins GitHubsite](https://github.com/CSSEGISandData/COVID-19). 
Link: https://github.com/CSSEGISandData/COVID-19

The agenda of this report is to analyze on below:

- Shows the cases and deaths due to COVID19 in all the countries since 2019.
- Show the number of cases and deaths due to COVID since 2019.
- Compare the fatality ratio in US verses all the countries in the world.

Case Fatality ratio is calculated as follows
$$ ( \frac{Number~of~Cases~reported~in~which~patient~died}{Number~of~Cases~Reported} ) * 100 $$

## Load Data

Below set of lines load the COVID19 data available at Johns Hopkins github site. It is the data about the confirmed cases and deaths are available in 5 separate csv file. 

To laod the data, I am saving the data to my local store from the John Hopkins  gitHub repository. If the file is not available in local store then i reload the file from the GitHub repository.

Each dataset contains the country/region, and the number of COVID-19 cases (or deaths or recoveries) reported daily.

```{r LoadData, include = TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_recovered_global.csv")
urls <- paste(url_in,file_names,sep = "")
urls

if(file.exists("global_confirmed_loaded.Rdata")) {
  print ('Loading Global Cases from cache')
  global_confirmed_loaded <- get(load("global_confirmed_loaded.Rdata"))
}else {
  print ('Loading Global Cases from URL')
  global_confirmed_loaded <- read_csv(urls[2])
  save(global_confirmed_loaded,file = "global_confirmed_loaded.Rdata")
}

global_confirmed_loaded

if(file.exists("global_death_loaded.Rdata")) {
  print ('Loading Global Deaths from cache')
  global_death_loaded <- get(load("global_death_loaded.Rdata"))
} else {
  global_death_loaded <- read_csv(urls[4])
  save(global_death_loaded,file = "global_death_loaded.Rdata")
}

if(file.exists("global_recovered_loaded.Rdata")) {
  global_recovered_loaded <- get(load("global_recovered_loaded.Rdata"))
} else {
  global_recovered_loaded <- read_csv(urls[5])
  save(global_recovered_loaded,file = "global_recovered_loaded.Rdata")
}

if(file.exists("us_confirmed_loaded.Rdata")) {
  print ('Loading US Confirmed from cache')
  us_confirmed_loaded <- get(load("us_confirmed_loaded.Rdata"))
} else {
  us_confirmed_loaded <- read_csv(urls[1])
  save(us_confirmed_loaded,file = "us_confirmed_loaded.Rdata")
}

if(file.exists("us_death_cache.Rdata")) {
  print ('Loading US Deaths from cache')
  us_death_loaded <-get(load("us_death_cache.Rdata"))
} else {
  us_death_loaded <- read_csv(urls[3])
  save(us_death_loaded,file = "us_death_cache.Rdata")
}

```

## Transforming Global COVID-19 data

The data containing the information about the COVID-19 cases from all the countries are available in different set such as confirmed COVID-19 cases per day per region and number of deaths due to COVID-19 per day per region.

```{r GlobalDataTransform, include = TRUE, echo = TRUE}

global_confirmed <- global_confirmed_loaded %>% 
  pivot_longer(
    cols  =-c('Province/State','Country/Region','Lat','Long'),
    names_to = "Date",
    values_to = "Cases") %>% 
  select(-c('Lat','Long'))

global_death <- global_death_loaded %>% 
  pivot_longer(
    cols = -c('Province/State','Country/Region','Lat','Long'),
    names_to = "Date",
    values_to = "Deaths") %>% 
  select(-c('Lat','Long'))

global_recovered <- global_recovered_loaded %>% 
  pivot_longer(cols = -c('Province/State','Country/Region','Lat','Long'),
               names_to = "Date",
               values_to = "Recovered") %>% 
  select(-c('Lat','Long'))

global_covid_data <- global_confirmed %>% 
  full_join(global_death) %>% 
  full_join(global_recovered) %>% 
  rename (
    Province_State = 'Province/State',
    Country_Region = 'Country/Region') %>% 
  mutate(Date = mdy(Date)) %>%
  select(-c(Province_State))

summary(global_covid_data )
global_covid_data

global_cases_by_month <- global_covid_data %>% 
  group_by( YearMonth = lubridate::floor_date(Date, 'month')) %>%
  drop_na() %>%
  summarize(Cases = sum(Cases), 
            Deaths = sum(Deaths),
            Recovered = sum(Recovered)) 


global_cases_by_month_w_fr <- global_cases_by_month %>% mutate(FatalityRatio = (Deaths/Cases * 100))
global_cases_by_month_w_fr


```
## Transforming the US COVID-19 data

The dataset containing the infrmation about the COVID-19 cases in US are transformed to get the number of confirmed cases and number of deaths due to COVID-19 per month. The data is also used to find the fatality rate of COVID-19 in US.

```{r DataTransform, include = TRUE, echo=TRUE, warning = FALSE}

us_confirmed <- us_confirmed_loaded %>% 
  pivot_longer(cols = -c("UID": 'Combined_Key'),
               names_to = "Date",
               values_to = "Cases") %>% 
  select('Admin2':'Cases') %>% 
  mutate(Date = mdy(Date)) %>% 
  select (-c('Lat', 'Long_')) %>% 
  rename ( County = 'Admin2')

us_death <- us_death_loaded %>% 
  pivot_longer(cols = -c("UID": 'Combined_Key'),
               names_to = "Date",
               values_to = "Deaths") %>% 
  select('Admin2':'Deaths') %>% 
  mutate(Date = mdy(Date)) %>% 
  select (-c('Lat', 'Long_')) %>% 
  rename ( County = 'Admin2')

us_cases <- us_confirmed %>% full_join(us_death) %>% filter(Cases > 0)
us_cases


us_cases_by_month <- us_cases %>% 
  group_by(YearMonth = lubridate::floor_date(Date, 'month'), Country_Region) %>%
  summarize(Cases = sum(Cases), 
            Deaths = sum(Deaths)) 


us_cases_by_month_w_fr <- us_cases_by_month %>% mutate(FatalityRatio = (Deaths/Cases * 100))
us_cases_by_month_w_fr

us_cases_by_state <- us_cases %>% 
  group_by(County, Province_State, Country_Region, Date) %>% 
  summarize(Total_Cases = sum(Cases), Total_Deaths = sum(Deaths)) %>% 
  select('Province_State', 'Country_Region', 'Date', 'Total_Cases', 'Total_Deaths') %>% 
  ungroup()

tail(us_cases_by_state)

us_total_cases <- us_cases_by_state %>% 
  group_by(Country_Region, Date) %>% 
  summarize( 
    Total_Cases = sum(Total_Cases), 
    Total_Deaths = sum(Total_Deaths)
    ) %>% 
  select(Country_Region, Date, Total_Cases, Total_Deaths) %>% 
  ungroup()

```

## Visualization

The below plot represents the COVID-19 trend for each year for the entire world. This visualization shows the number of COVID-19 cases and deaths due to COVID-19 happened yearly.


```{r DataVisualizationWorld, include = TRUE, echo = TRUE}
 ggplot(data = global_cases_by_month, aes(x=YearMonth,y=Cases)) + 
  geom_line(aes(color='Cases')) + 
  geom_point(aes(color='Cases')) + 
  geom_point(aes(y= Deaths, color='Deaths')) +
  geom_line(aes(y=Deaths, color = 'Deaths')) +
  scale_y_log10() +
  #scale_y_continuous(labels = scales::label_number_si()) +
  #scale_y_continuous(labels = scales::scale_y_log10()) +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in World", y = 'Cases', x = 'Year')

```


The below plot represents the COVID-19 trend for each year for United States. This visualization shows the number of COVID-19 cases and deaths due to COVID-19 yearly.

```{r DataVisualizationUS, include = TRUE, echo = TRUE}

addUnits <- function(n) {
  labels <- ifelse(n < 1000, n,  # less than thousands
                   ifelse(n < 1e6, paste0(round(n/1e3), 'k'),  # in thousands
                          ifelse(n < 1e9, paste0(round(n/1e6), 'M'),  # in millions
                                 ifelse(n < 1e12, paste0(round(n/1e9), 'B'), # in billions
                                        ifelse(n < 1e15, paste0(round(n/1e12), 'T'), # in trillions
                                               'too big!'
                                        )))))
  return(labels)
}

ggplot(data = us_cases_by_month, aes(x=YearMonth,y=Cases)) + 
  geom_line(aes(color='Cases')) + 
  geom_point(aes(color='Cases')) + 
  geom_point(aes(y= Deaths, color='Deaths')) +
  geom_line(aes(y=Deaths, color = 'Deaths')) +
  scale_y_log10() +
  #scale_y_continuous(labels = addUnits)
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US",y = 'Cases',  x = 'Year')

```

The below plot represents the comparison of the fatality rate globally verse the United States yearly. The fatality is the number of deaths in the reported COVID cases per number of COVID-19 cases reported. 

```{r DataVisualizationFatality, include = TRUE, echo = TRUE}
all_fatality_rate <- us_cases_by_month_w_fr %>% 
  rename(FatalityRatioUS = FatalityRatio) %>%
  right_join(global_cases_by_month_w_fr,by="YearMonth") %>%
  rename(FatalityRatioAllCountries = FatalityRatio) %>%
  select (c(YearMonth, FatalityRatioUS, FatalityRatioAllCountries))

ggplot(data = all_fatality_rate, aes(x=YearMonth,y=FatalityRatioUS)) + 
  geom_point(aes(color='US')) + 
  geom_line(aes(color='US')) + 
  geom_point(aes(y= FatalityRatioAllCountries, color='AllCountries')) +
  geom_line(aes(y=FatalityRatioAllCountries, color = 'AllCountries')) +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = "Fatality Rate", x='Year', y= 'Fatality Rate')


```

## Modelling

The below model identifies the number of deaths based on the total number of cases reported. 

**Please note that it doesn't consider external factors like availability of vaccination, immunity gained in people who already had COVID in the past, etc. The Model uses all the available data to train the data, and the same data is used to plot the values to check how well the model is trained on the current data. Ideally, a different dataset should have been used to test it well.**
```{r Model, echo=TRUE}

mod <- lm(Total_Deaths ~ Total_Cases, data = us_total_cases)
summary(mod)

us_total_cases_w_pred <- us_total_cases %>% mutate(PRED_DEATHS = predict(mod))
 
ggplot(data = us_total_cases_w_pred ) + 
  geom_point(aes(x = Total_Cases, y = Total_Deaths ), color = "blue") +
  geom_point(aes(x = Total_Cases, y = PRED_DEATHS ), color = "red") +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle=90)) +
  labs(title = "COVID-19 Actual verses Predicted", x='Cases', y= 'Deaths')
```

## Bias

The data is available on John Hopkins Github and is provided by respective countries' administrations. The facilities in each country are different due to different factors such as population, people willing to get the COVID test, the socio-economic condition in that country, etc. 
Countries that doesn't ave enough facilities might not be able to test the COVID as mush as other developed countries. This factor could lead to bias in the analysis. 

## Summary
The reported cases in the USA were high compared to other countries based on the population. But based on the analysis, the trend of COVID cases in the US compared to the world trend was very similar. Compared to the other factors, such as the fatality rate between the same parties, the trend again looked very similar. It might be possible that the US has more cases reported since the US is developedoped countries and has better facilities to test, track and monitor the COVID cases.